Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-04 scagnostics.Rmd
Moving Figures to appendix, remove some commented old text.

<!--
We first study the smoothness of each measure calculated on a sequence of 2-d projections obtained via an interpolated planned tour path moving between the projection onto $x_1$-$x_2$ to the projection onto $x_3$-$x_4$ and using the default interpolation angle of 0.05. Each time step refers to an interpolation step. Since the value of the TIC index is strongly dependent on the input distribution sample size we have normalised the TIC measure by its maximum for each individual dataset, all other measures are reproduced as reported by the calculation. The results are shown in Figure \ref{fig:plotV1V2toV3V4}. The standard index functions "holes" and "cmass" show the desired behaviour in that the change is smooth over interpolated views. Considering now the indices from the scagnostics family, we first note stronge dependence on the sample size. The convex index is largely sensitive to random fluctuations when considering small samples ($n=100$), while being mostly flat for larger samples, especially in the case of an underlying uniform distribution. The skinny index on the other hand is systematically larger for the smaller samples, while stringy appears not to be sensitive to the sample size, but resulting in large fluctuations and sharp jumps for all datasets. Distance correlation and the spline based index show similar behaviour, i.e. when considering large enough data samples ($n=1000$) the values are consistently close to zero, as expected for a distribution without structure. On the other hand somewhat larger values (but still far from the maximum value of 1) are found when considering small datasets, where distributions are however mostly smooth, with few kinks depending on index function and underlying distribution. Finally the information based metrics also show systematic differences based on the sample size, with sampling fluctuations being significant for small datasets. Note here that fluctuations present in Figure \ref{fig:plotV1V2toV3V4} may be amplified, as maximum values for most index functions are far from the theoretic maximum.
-->

@@

<!--
```{r plannedtourpath}
if(!file.exists("cache/plannedtourpath.rda")){
  m1 <- matrix(c(1,0,0,0,0,0,0,1,0,0,0,0),ncol=2)
  m2 <- matrix(c(0,0,0,0,1,0,0,0,0,0,0,1),ncol=2)
  #this is silly but seems that first two entries are being ignored, so need some fake entries
  m3 <- matrix(c(1,0,0,0,1,0,0,1,0,0,0,1),ncol=2)
  m4 <- matrix(c(1,1,1,0,0,0,0,1,1,0,0,0),ncol=2)
  t2 <- save_history(sin100,tour_path=planned_tour(list(m3,m4,m1,m2)))
  t2full <- as.list(interpolate(t2))
  fullResPlanned <- getProj(sin100, t2full, "sin", 100) %>%
    rbind(getProj(sin1000, t2full, "sin", 1000)) %>%
    rbind(getProj(spiral100, t2full, "spiral", 100)) %>%
    rbind(getProj(spiral1000, t2full, "spiral", 1000)) %>%
    rbind(getProj(pipe100, t2full, "pipe", 100)) %>%
    rbind(getProj(pipe1000, t2full, "pipe", 1000))
  save(t2, t2full, fullResPlanned, file = "cache/plannedtourpath.rda")
} else {
  load("cache/plannedtourpath.rda")
}
```

```{r plannedtour, fig.width=6, fig.height=7, fig.cap="PPIs for projections along an interpolation between a nuisance and a structured projection (solid lines: $n=1000$; dashed: $n=100$). Most of the indexes have peaks at the end of the sequence, indicating that they see the structure. The scagnostics, MIC and TIC see all three structures, so are more flexible for general pattern detection."}
fullResMeltPlanned <- fullResPlanned %>%
  mutate(convex = 1-convex) %>% 
  rowwise() %>%
  mutate(holes = rescaleHoles(holes)) %>%
  gather(PPI, value, -t, -name, -size) %>%
  filter(PPI != "cmass") %>%
  mutate(PPI = fct_relevel(PPI, "holes", "convex", "skinny", "stringy", "dcor2d", "splines2d", "MIC", "TIC"))
fullResMeltPlanned %>% 
  ggplot(aes(x=t, y=value)) + 
  geom_line(aes(group=size, linetype=factor(size, levels=c("1000", "100")))) + 
  ylim(c(0,1)) +
  facet_grid(PPI~name) + 
  theme(legend.position="none") + xlab("Sequence of projections (t)") + ylab("PPI value")
```

## Tentative
try how this looks going first to V1 vs V5, then to V5 vs V6
-->

@@

The optimization is done in two stages, a scouting phase using method 2, and a refinement stage using method 3.  For the scouting we use $\alpha = 0.5$ and stopping condition of maximum 5000 tries, and we optimist the TIC index. The results are shown in Figures \ref{fig:pipeFirstRun} and \ref{fig:testpipe}.

@@

Figure \ref{fig:testpipe} shows the projection returned during the scouting phase (left) and the refinement phase (right). It was important to start the method 3 optimizer at the best projection returned during the scouting phase, to smoothly converge more closely to the ideal projection.

@@

Figure \ref{fig:neutronStarSPLOM} shows the scatterplot matrix (with transparency) of the six parameters. (In astrophysics, scatterplot matrices are often called "corner plots" [@corner].) The diagonal shows a univariate density plot of each parameter, and the upper triangle of cells displays the correlation between variables. From this it can be seen that m1 and m2 are strongly, and slightly, nonlinearly associated. Between the other variables, some linear association (R1, R2), some nonlinear association (L1, L2, R1, R2),  heteroskedastic variance in most variables and some bimodality (R1, L1, L2, m1, m2).

@@

<!--
(left) clearly recovers functional relations between some of the parameters. From the matrix representation we read off that parameters m1, L1 and R1 are important in this projection. Using the original (non-rescaled) data and taking into account the difference in scale between m1 and R1 we can reproduce a similar picture shown in Fig \ref{fig:nsePlotOrig} (middle), which we compare to the dependence of L1 on the ratio R1/m1 (right). For this example the linear combination can thus recover the approximate relation between these parameters.
-->

<!--
 A posterior sample containing 2538 points is available @ligoData and can be used to constrain the parameters. Here we use the "Parametrized-EoS\_nomaxmass\_posterior\_samples.dat" dataset and aim to recover the dependence between the three parameters of the better constrained system out of a larger parameter set. For details about the modelling and statistical procedure see @Abbott2018exr. A common representation of posterior samples used for interpretation are so-called "corner plots", i.e. density displays for all 2-d parameter combinations as well as the 1-d profile, highlighting empirically evaluated regions of highest posterior density, which correspond to credibility regions at a given level. The information can similarly be presented in a scatter plot matrix using transparency to convey the density information. Since in the following we work with 2-d scatter plots we choose to show the data in this fashion, see Fig. \ref{fig:neutronStarSPLOM}.
-->

```{r neutronStarSPLOM, fig.height=8, fig.width=8, fig.cap="Scatter plot matrix of the neutron star dataset, darker regions represent higher marginalised posterior densities."}
nsD <- read_csv("data/samples.csv") 
ggpairs(nsD, lower=list(continuous = wrap("points", alpha = 0.05)))
```

<!-- The following features can be observed: the combination of the two mass parameters is well constrained (a consequence of how a particular combination, the "chirp mass" enters predictions), other parameters also show patterns of dependence, they are however less pronounces. We also see the non-linear relation between the radius $R_1$ and $\Lambda$ parameter $L1$ and the two-pronged disribution for the radius $R_1$.
For further study we therefore remove the second mass $m2$ from the dataset.
\textcolor{red}{Looking at the distribution in a grand tour display, using a scatter plot display combined with alpha transparency to study the density distribution in the 5-d parameter space, indicates that the data points are found on a curved surface in the five dimensional space. This stems from the non-linear relation between the parameters, which we will try to uncover using the above developed formalism for new guided tour index functions.}

### Results

Since we expect a functional relation between some of the parameters, we optimise the "splines2d" index function. Note that all parameters are standardised to the range $[0,1]$ before running the optimisation, and we do not sphere the data for ease of interpretation of the results.
-->

```{r neutronStarEq,  results="hide"}
nsM <- nsD %>%
  select(-m2) %>% # remove m2 to avoid finding well known correlation
  rescale() # rescaling, this returns a matrix
if(!file.exists("cache/neutronStarEq.rda")){
  set.seed(2018)
  nseTour <- save_history(nsM, guided_tour(splineIndex()))
  nseTourFull <- as.list(interpolate(nseTour))
  save(nseTour, nseTourFull, file = "cache/neutronStarEq.rda")
} else {
  load("cache/neutronStarEq.rda")
}
iLast <- length(nseTourFull)
fProj <- nseTourFull[[iLast]]
dProj <- as.tibble(nsM %*% fProj)
colnames(fProj) <- c("PP1", "PP2")
colnames(dProj) <- c("PP1", "PP2")
fProj <- as_tibble(fProj) %>%
  add_column(label=colnames(nsM))
```

@@

Figure \ref{fig:bbhSimulation} shows a scatterplot matrix, of nine of the twelve parameters. (Parameter m2 is not shown because it is strongly linearly associated with m1, phi$\_$jl and psi are not shown because they uniform, and not associated with other parameters.) Among the nine plotted parameters, strong nonlinear relationships can be seen between the parameters ra, dec and time. The first two describe the position of the event in the sky, and time is the merging time (in GPS units). Because of the elliptical relationship between dec and time, the TIC index is used for PP, even though it is slow to compute. Between the other parameters, the main structure seen is multimodality and some skewness. These patterns are representative of the likelihood function, since most priors are flat, or built to capture growth with volume rather than distance.

```{r bbhSimulation, fig.height=8, fig.width=8, dev = "png", dpi=300, fig.cap="Scatter plot matrix showing most of the variables included in the BBH dataset. Strong correlation between the parameters time, dec and ra can be observed."}
bbhD <- read_csv("data/posterior_samples.csv")
bbhDsmall <- select(bbhD, -phi_jl, -m2, -psi, -chi_eff)
ggpairs(bbhDsmall, lower=list(continuous = wrap("points", alpha = 0.02)))
```

@@

```{r bbhGuided, results="hide", fig.cap="Projections corresponding to the maxima of three indices: TIC, splines2D and 1-convex. Projections a, b found by TIC and splines3D are very similar, and involve the same three parameters, ra, dec and time. The 1-convex index finds a very different view.", dev = "png", dpi=300}
bbhM <- rescale(select(bbhD,-m2, -chi_eff))
if(!file.exists("cache/bbhGuidedTIC.rda")){
  set.seed(2018)
  bbhTour1 <- save_history(bbhM, guided_tour(mineIndex("TIC")))
  bbhTourFull1 <- as.list(interpolate(bbhTour1))
  save(bbhTour1, bbhTourFull1, file = "cache/bbhGuidedTIC.rda")
} else {
  load("cache/bbhGuidedTIC.rda")
}
if(!file.exists("cache/bbhGuidedSpline.rda")){
  set.seed(2018)
  bbhTour2 <- save_history(bbhM, guided_tour(splineIndex()))
  bbhTourFull2 <- as.list(interpolate(bbhTour2))
  save(bbhTour2, bbhTourFull2, file = "cache/bbhGuidedSpline.rda")
} else {
  load("cache/bbhGuidedSpline.rda")
}
if(!file.exists("cache/bbhGuidedConvex.rda")){
  set.seed(2018)
  bbhTour3 <- save_history(bbhM, guided_tour(invConvexIndex()))
  bbhTourFull3 <- as.list(interpolate(bbhTour3))
  save(bbhTour3, bbhTourFull3, file = "cache/bbhGuidedConvex.rda")
} else {
  load("cache/bbhGuidedConvex.rda")
}

@@

## Re-scaling of holes index {-}

@@

>>>>

@@

@@

The optimization is done in two stages, a scouting phase using method 2, and a refinement stage using method 3.  For the scouting we use $\alpha = 0.5$ and stopping condition of maximum 5000 tries, and we optimist the TIC index. The results are shown in Figures \ref{fig:pipeFirstRun}.

@@

The final views obtained in each of the two stages are compared in the Appendix, see Fig. \ref{fig:testpipe}.

@@

Figure \ref{fig:neutronStarSPLOM} in the Appendix shows the scatterplot matrix (with transparency) of the six parameters. (In astrophysics, scatterplot matrices are often called "corner plots" [@corner].) The diagonal shows a univariate density plot of each parameter, and the upper triangle of cells displays the correlation between variables. From this it can be seen that m1 and m2 are strongly, and slightly, nonlinearly associated. Between the other variables, some linear association (R1, R2), some nonlinear association (L1, L2, R1, R2),  heteroskedastic variance in most variables and some bimodality (R1, L1, L2, m1, m2).

@@

```{r neutronStarEq,  results="hide"}
nsD <- read_csv("data/samples.csv") 
nsM <- nsD %>%
  select(-m2) %>% # remove m2 to avoid finding well known correlation
  rescale() # rescaling, this returns a matrix
if(!file.exists("cache/neutronStarEq.rda")){
  set.seed(2018)
  nseTour <- save_history(nsM, guided_tour(splineIndex()))
  nseTourFull <- as.list(interpolate(nseTour))
  save(nseTour, nseTourFull, file = "cache/neutronStarEq.rda")
} else {
  load("cache/neutronStarEq.rda")
}
iLast <- length(nseTourFull)
fProj <- nseTourFull[[iLast]]
dProj <- as.tibble(nsM %*% fProj)
colnames(fProj) <- c("PP1", "PP2")
colnames(dProj) <- c("PP1", "PP2")
fProj <- as_tibble(fProj) %>%
  add_column(label=colnames(nsM))
```

@@

Figure \ref{fig:bbhSimulation} in the Appendix shows a scatterplot matrix, of nine of the twelve parameters. (Parameter m2 is not shown because it is strongly linearly associated with m1, phi$\_$jl and psi are not shown because they uniform, and not associated with other parameters.) Among the nine plotted parameters, strong nonlinear relationships can be seen between the parameters ra, dec and time. The first two describe the position of the event in the sky, and time is the merging time (in GPS units). Because of the elliptical relationship between dec and time, the TIC index is used for PP, even though it is slow to compute. Between the other parameters, the main structure seen is multimodality and some skewness. These patterns are representative of the likelihood function, since most priors are flat, or built to capture growth with volume rather than distance.

@@

```{r bbhGuided, results="hide", fig.cap="Projections corresponding to the maxima of three indices: TIC, splines2D and 1-convex. Projections a, b found by TIC and splines3D are very similar, and involve the same three parameters, ra, dec and time. The 1-convex index finds a very different view.", dev = "png", dpi=300}
bbhD <- read_csv("data/posterior_samples.csv")
bbhM <- rescale(select(bbhD,-m2, -chi_eff))
if(!file.exists("cache/bbhGuidedTIC.rda")){
  set.seed(2018)
  bbhTour1 <- save_history(bbhM, guided_tour(mineIndex("TIC")))
  bbhTourFull1 <- as.list(interpolate(bbhTour1))
  save(bbhTour1, bbhTourFull1, file = "cache/bbhGuidedTIC.rda")
} else {
  load("cache/bbhGuidedTIC.rda")
}
if(!file.exists("cache/bbhGuidedSpline.rda")){
  set.seed(2018)
  bbhTour2 <- save_history(bbhM, guided_tour(splineIndex()))
  bbhTourFull2 <- as.list(interpolate(bbhTour2))
  save(bbhTour2, bbhTourFull2, file = "cache/bbhGuidedSpline.rda")
} else {
  load("cache/bbhGuidedSpline.rda")
}
if(!file.exists("cache/bbhGuidedConvex.rda")){
  set.seed(2018)
  bbhTour3 <- save_history(bbhM, guided_tour(invConvexIndex()))
  bbhTourFull3 <- as.list(interpolate(bbhTour3))
  save(bbhTour3, bbhTourFull3, file = "cache/bbhGuidedConvex.rda")
} else {
  load("cache/bbhGuidedConvex.rda")
}

@@

## Re-scaling of holes index

@@

## Additional Figures

### Final projection of pipe guided tour
Figure \ref{fig:testpipe} shows the projection returned during the scouting phase (left) and the refinement phase (right). It was important to start the method 3 optimizer at the best projection returned during the scouting phase, to smoothly converge more closely to the ideal projection.

### Dataset overview scatter plot matrices

```{r neutronStarSPLOM, fig.height=8, fig.width=8, fig.cap="Scatter plot matrix of the neutron star dataset, darker regions represent higher marginalised posterior densities."}
ggpairs(nsD, lower=list(continuous = wrap("points", alpha = 0.05)))
```

```{r bbhSimulation, fig.height=8, fig.width=8, dev = "png", dpi=300, fig.cap="Scatter plot matrix showing most of the variables included in the BBH dataset. Strong correlation between the parameters time, dec and ra can be observed."}
bbhDsmall <- select(bbhD, -phi_jl, -m2, -psi, -chi_eff)
ggpairs(bbhDsmall, lower=list(continuous = wrap("points", alpha = 0.02)))
```
